Forecasting the Support of Kamala Harris in 2024 U.S. Presidential Election using a Model-Based Approach

### 2024-11-04 paper/ForecastingtheSupportofKamalaHarrisin2024U.S.PresidentialElection.qmd
update paper
update typo/grammar error

---
title: "Forecasting the Support of Kamala Harris in 2024 U.S. Presidential Election using a Model-Based Approach"
subtitle: "Kamala Harris expects to win on 47.5%, leads Donald Trump by about 1% in 2024 US election in Popular Vote based on Poll-of-Polls Data and Linear Modeling."
author: 
  - Diana Shen
  - Jinyan Wei
  - Jerry Yu
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Forecast-2024-US-election](https://github.com/DianaShen1224/Forecast-2024-US-election)."
date: today
date-format: long
abstract: "This study models the 2024 U.S. Presidential Election by analyzing polling data with adjustments for factors like recency and sample size, simulating an approach that captures up-to-date voter sentiment. Using the recent 4 months (July 21th - November 2th) of aggregated poll data, we built a multiple linear regression model to estimate Harris’ support in popular vote and the winner of seven key battleground states. We estimate that  Kamala Harris will lead with an expected 47.5% support, ahead of Donald Trump by approximately 1% in popular vote and winning in three of seven key battleground states, Wisconsin, Michigan, and Pennsylvania.  By emphasizing recent, high-quality polls, the analysis improves predictive reliability, offering an understanding that could inform campaign strategies and public discourse. These findings highlight the value of weighting and time-sensitive adjustments in forecasting methods, balancing accuracy with responsiveness to the latest trends."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

In this study, we utilized several R packages to enhance our data
manipulation, modeling, and visualization capabilities. The `tidyverse`
package offered a detailed set of tools for data wrangling and
analysis, improving workflow efficiency [@thereferencecanbewhatever].
The `here` package aided in managing file paths, improving the reproducible workflow of our study [@citehere].We use `testthat` to verify dataset variables, ensuring they are correctly formatted and contain expected values for analysis [@citetestthat]. We relied on `janitor` to perform data
cleaning, as it provides functionalities to identify and rectify quality
issues within the dataset [@citejanitor]. For handling date-related
operations, the `lubridate` package help simplifying the
manipulation of time variables [@citelubridate]. `arrow` supported
efficient data input and output in a parquet format,
essential for managing larger datasets [@citearrow]. We use `ggplot2` [@citeggplot2] to visualize the analysis of data. In addition to the core packages, we employed several specialized libraries to facilitate specific analytical tasks. The `modelsummary` package enabled us to generate clean, interpretable model summaries, streamlining the reporting of regression and other statistical results [@citemodelsummary]. `RColorBrewer` provided color palettes for visualizations, enhancing clarity and visual appeal in our plots [@citeRColorBrewer]. The `future` package supported parallel processing, optimizing the runtime of intensive computations, which is especially beneficial in model training [@citefuture]. Finally,for validation of model using K-Fold cross validation, we utilized the `boot` package, which allowed us to assess model stability and reliability [@citeboot1;@citeboot2] and `caret` to facilitate model training and validation, offering a unified interface for cross-validation and performance metrics [@citecaret]. Our coding practices and file organization were informed by the structure outlined
in @tellingstories.

@@

### Recency {#sec-recency}
```{r}
#| label: fig-recency
#| fig-cap: The distribution of the Poll Recency since July 21, 2024 highlights a strong focus on recent polling data, with the highest frequency of polls occurring within the first few days.
#| echo: false
#| warning: false
#| message: false
ggplot(harris_data, aes(x = recency)) +
  geom_histogram(binwidth = 1, fill = "gold", color = "grey50") +
  labs(
    x = "Recency (days)",
    y = "Count"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

@@

@fig-numericgrade displays the distribution of numeric grades assigned to various pollsters. The numeric grade is a metric that evaluates the quality or reliability of a pollster, considering factors such as methodology, historical accuracy, sample size, and pollster reputation. In this chart, we observe that the frequency of numeric grades peaks around 2.75 and 3.00, with a notable concentration at these values. This suggests that, after excluding lower-quality polls, a significant number of surveys are conducted by highly rated pollsters. Fewer polls have grades below 2.5, indicating a relatively lower occurrence of polls from less-reliable pollsters in this dataset. This distribution highlights the emphasis on high-quality pollsters, ensuring a more reliable and consistent source of polling data for subsequent analysis.

```{r}
#| label: fig-numericgrade
#| fig-cap: The distribution of the Numeric Grade of the Polls shows that surveys are conducted by highly rated pollsters, which ensure the reliability of the analysis.
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: fig-percentage
#| fig-cap: The distribution of the Support Percentage of Kamala Harris clustered around 50%, with the majority of support falling between 40% and 55%.
#| echo: false
#| warning: false
#| message: false
ggplot(harris_data, aes(x = pct)) +
  geom_histogram(binwidth = 1, fill = "gold", color = "grey50") +
  labs(
    x = "Support for Harris (%)",
    y = "Count"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

@@

In this analysis, we use predictors that capture both the methodological
quality and structural aspects of each poll. Specifically, we include
variables such as `national_poll`, which reflects national support
trends; `pollster`, which represents differences in methodology and
reliability across polling organizations; `population`, which
accounts for the type of respondent group of voters associated with
each poll; and lastly,`state`, which represents the exact U.S. state
where the poll was conducted or focused.Our combined weighting approach further
integrates factors like recency and sample size to ensure that more recent, 
larger-sample polls have a greater impact on the model's predictions, providing a
balanced and detailed assessment of candidate support.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| tbl-cap: "Summary of Multiple Linear Regression Model Predicting Harris Support Based on Polling Data since 21 July 2024 (Harris first declared) to 2 November 2024"
#| label: tbl-summary1
msummary(
  list("Unweighted" = harris_unweighted_model, "Weighted" = harris_weighted_model),
  fmt = fmt_significant(3),coef_omit="state|pollster",notes = "Note. All models include a control for state and pollsters. The reference level of state is Alaska, the reference level of population is likely voter, and the reference level of pollsters is Angus Reid.")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| tbl-cap: "Summary of Multiple Linear Regression Model Predicting Trump Support Based on Polling Data since 21 July 2024 (Harris first declared) to 2 November 2024"
#| label: tbl-summary2
msummary(
  list("Unweighted" = trump_unweighted_model, "Weighted" = trump_weighted_model),
  fmt = fmt_significant(3),coef_omit="state|pollster",notes = "Note. All models include a control for state and pollsters. The reference level of state is Alaska,the reference level of population is likely voter, and the reference level of pollsters is ABC/Washington Post.")
```

@@

set.seed(233)
plan(multisession)
####### Prediction for Harris ######
new_data_harris <- data.frame(
  end_date = seq(
    min(harris_data$end_date),
    max(harris_data$end_date),
    length.out = 700
  ),
  state = factor(sample( c("National","Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin"), size = 700, replace = TRUE)), 
  pollster = factor(sample(levels(harris_data$pollster), size = 700, replace = TRUE), 
                    levels = levels(harris_data$pollster)), 
  population = factor(sample(levels(harris_data$population), size = 700, replace = TRUE), 
                    levels = levels(harris_data$population)),
  sample_size = sample(1:2300, size = 700, replace = TRUE)
)|>
  mutate(
    national_poll = factor(if_else(state == "National", 1, 0)),
    recency = as.numeric(difftime(as.Date("2024-11-05"), end_date, units = "days")),
    sample_size_weight = pmin(sample_size / 2300, 1),
    recency_weight = exp(-recency * 0.1),
    combined_weight = recency_weight * sample_size_weight,
    candidate="Kamala Harris"
  )

####### Prediction for Trump ######
new_data_trump <- data.frame(
  end_date = seq(
    min(trump_data$end_date),
    max(trump_data$end_date),
    length.out = 700
  ),
  pollster = factor(sample(levels(trump_data$pollster), size = 700, replace = TRUE), 
                    levels = levels(trump_data$pollster)),
 population= factor(sample(levels(trump_data$population), size = 700, replace = TRUE), 
                    levels = levels(trump_data$population)),
   state = factor(sample( c("National","Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin"), size = 700, replace = TRUE), 
                    levels = levels(trump_data$state)),  
  sample_size =sample(1:2300, size = 700, replace = TRUE)
)|>
  mutate(
   national_poll = factor(if_else(state == "National", 1, 0)),
      recency = as.numeric(difftime(as.Date("2024-11-05"), end_date, units = "days")),
    sample_size_weight = pmin(sample_size / 2300, 1),
    recency_weight = exp(-recency * 0.1),  
    combined_weight = recency_weight * sample_size_weight,
    candidate="Donald Trump"
  )
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-cap: "Harris Kamala leads Donald Trump an average 1% in Predicted Support in the 2024 US Election (From August 1, 2024, to November 2, 2024)."
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| label: fig-support-trend
# Plotting with the filtered data
plan(multisession)
##unweighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),size = 0.2)+
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",caption = 
                   "Model: Candidate Support Percentage = national_poll + pollster + population + state") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)

>>>>

---
title: "Forecasting the Support of Kamala Harris in 2024 U.S. Presidential Election using a Model-Based Approach"
subtitle: "Kamala Harris expects to win on 47.5%, leads Donald Trump by about 1% in 2024 US election in Popular Vote based on Poll-of-Polls Data and Linear Modeling."
author: 
  - Diana Shen
  - Jinyan Wei
  - Jerry Yu
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Forecast-2024-US-election](https://github.com/DianaShen1224/Forecast-2024-US-election)."
date: today
date-format: long
abstract: "This study models the 2024 U.S. Presidential Election by analyzing polling data with adjustments for factors like recency and sample size, simulating an approach that captures up-to-date voter sentiment. Using the recent 4 months (July 21st - November 2nd) of aggregated poll data, we built a multiple linear regression model to estimate Harris’ support in the popular vote and the winner of seven key battleground states. We estimate that  Kamala Harris will lead with an expected 47.5% support, ahead of Donald Trump by approximately 1% in the popular vote and winning in three of seven key battleground states, Wisconsin, Michigan, and Pennsylvania.  By emphasizing recent, high-quality polls, the analysis improves predictive reliability, offering an understanding that could inform campaign strategies and public discourse. These findings highlight the value of weighting and time-sensitive adjustments in forecasting methods, balancing accuracy with responsiveness to the latest trends."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

In this study, we utilized several R packages to enhance our data
manipulation, modeling, and visualization capabilities. The `tidyverse`
package offered a detailed set of tools for data wrangling and
analysis, improving workflow efficiency [@thereferencecanbewhatever].
The `here` package aided in managing file paths, improving the reproducible workflow of our study [@citehere].We use `testthat` to verify dataset variables, ensuring they are correctly formatted and contain expected values for analysis [@citetestthat]. We use `knitr` to build the table using the `kable` function [@citeknitr]. We relied on `janitor` to perform data
cleaning, as it provides functionalities to identify and rectify quality
issues within the dataset [@citejanitor]. For handling date-related
operations, the `lubridate` package helps simplify the
manipulation of time variables [@citelubridate]. `arrow` supported
efficient data input and output in a parquet format,
essential for managing larger datasets [@citearrow]. We use `ggplot2` [@citeggplot2] to visualize the analysis of data. In addition to the core packages, we employed several specialized libraries to facilitate specific analytical tasks. The `modelsummary` package enabled us to generate clean, interpretable model summaries, streamlining the reporting of regression and other statistical results [@citemodelsummary]. `RColorBrewer` provided color palettes for visualizations, enhancing clarity and visual appeal in our plots [@citeRColorBrewer]. The `future` package supported parallel processing, optimizing the runtime of intensive computations, which is especially beneficial in model training [@citefuture]. Finally, for validation of the model using K-Fold cross-validation, we utilized the `boot` package, which allowed us to assess model stability and reliability [@citeboot1;@citeboot2] and `caret` to facilitate model training and validation, offering a unified interface for cross-validation and performance metrics [@citecaret]. Our coding practices and file organization were informed by the structure outlined
in @tellingstories.

@@

### Recency {#sec-recency}
```{r}
#| label: fig-recency
#| fig-cap: The distribution of the Poll Recency since July 21, 2024, highlights a strong focus on recent polling data, with the highest frequency of polls occurring within the first few days.
#| echo: false
#| warning: false
#| message: false
ggplot(harris_data, aes(x = recency)) +
  geom_histogram(binwidth = 1, fill = "gold", color = "grey50") +
  labs(
    x = "Recency (days)",
    y = "Count"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

@@

@fig-numericgrade displays the distribution of numeric grades assigned to various pollsters. The numeric grade is a metric that evaluates the quality or reliability of a pollster, considering factors such as methodology, historical accuracy, sample size, and pollster reputation. In this chart, we observe that the frequency of numeric grades peaks around 2.75 and 3.00, with a notable concentration at these values. This suggests that, after excluding lower-quality polls, a significant number of surveys are conducted by highly rated pollsters. Fewer polls have grades below 2.5, indicating a relatively lower occurrence of polls from less reliable pollsters in this dataset. This distribution highlights the emphasis on high-quality pollsters, ensuring a more reliable and consistent source of polling data for subsequent analysis.

```{r}
#| label: fig-numericgrade
#| fig-cap: The distribution of the Numeric Grade of the Polls shows that surveys are conducted by highly rated pollsters, which ensures the reliability of the analysis.
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: fig-percentage
#| fig-cap: The distribution of Kamala Harris's support percentage clustered around 50%, with the majority of support falling between 40% and 55%.
#| echo: false
#| warning: false
#| message: false
ggplot(harris_data, aes(x = pct)) +
  geom_histogram(binwidth = 1, fill = "gold", color = "grey50") +
  labs(
    x = "Support for Harris (%)",
    y = "Count"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

@@

In this analysis, we use predictors that capture both the methodological
quality and structural aspects of each poll. Specifically, we include
variables such as `national_poll`, which reflects national support
trends; `pollster`, which represents differences in methodology and
reliability across polling organizations; `population`, which
accounts for the type of respondent group of voters associated with
each poll; and lastly, `state`, which represents the exact U.S. state
where the poll was conducted or focused.Our combined weighting approach further
integrates factors like recency and sample size to ensure that more recent, 
larger-sample polls have a greater impact on the model's predictions, providing a
balanced and detailed assessment of candidate support.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| tbl-cap: "Summary of Multiple Linear Regression Model Predicting Harris Support Based on Polling Data since 21 July 2024 (Harris first declared) to 2 November 2024"
#| label: tbl-summary1
msummary(
  list("Unweighted" = harris_unweighted_model, "Weighted" = harris_weighted_model),
  fmt = fmt_significant(3),coef_omit="state|pollster",notes = "Note. All models include a control for state and pollsters. The reference level of the state is Alaska, the reference level of population is likely voter, and the reference level of pollsters is Angus Reid.")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| tbl-cap: "Summary of Multiple Linear Regression Model Predicting Trump Support Based on Polling Data since 21 July 2024 (Harris first declared) to 2 November 2024"
#| label: tbl-summary2
msummary(
  list("Unweighted" = trump_unweighted_model, "Weighted" = trump_weighted_model),
  fmt = fmt_significant(3),coef_omit="state|pollster",notes = "Note. All models include a control for state and pollsters. The reference level of the state is Alaska, the reference level of population is likely voter, and the reference level of pollsters is ABC/Washington Post.")
```

@@

set.seed(233)
plan(multisession)
####### Prediction for Harris ######
new_data_harris <- data.frame(
  end_date = seq(
    min(harris_data$end_date),
    max(harris_data$end_date),
    length.out = 700
  ),
  state = factor(sample( c("National", "Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin"), size = 700, replace = TRUE)), 
  pollster = factor(sample(levels(harris_data$pollster), size = 700, replace = TRUE), 
                    levels = levels(harris_data$pollster)), 
  population = factor(sample(levels(harris_data$population), size = 700, replace = TRUE), 
                    levels = levels(harris_data$population)),
  sample_size = sample(1:2300, size = 700, replace = TRUE)
)|>
  mutate(
    national_poll = factor(if_else(state == "National", 1, 0)),
    recency = as.numeric(difftime(as.Date("2024-11-05"), end_date, units = "days")),
    sample_size_weight = pmin(sample_size / 2300, 1),
    recency_weight = exp(-recency * 0.1),
    combined_weight = recency_weight * sample_size_weight,
    candidate="Kamala Harris"
  )

####### Prediction for Trump ######
new_data_trump <- data.frame(
  end_date = seq(
    min(trump_data$end_date),
    max(trump_data$end_date),
    length.out = 700
  ),
  pollster = factor(sample(levels(trump_data$pollster), size = 700, replace = TRUE), 
                    levels = levels(trump_data$pollster)),
 population= factor(sample(levels(trump_data$population), size = 700, replace = TRUE), 
                    levels = levels(trump_data$population)),
   state = factor(sample( c("National", "Arizona", "Georgia", "Michigan", "Nevada", "North Carolina", "Pennsylvania", "Wisconsin"), size = 700, replace = TRUE), 
                    levels = levels(trump_data$state)),  
  sample_size =sample(1:2300, size = 700, replace = TRUE)
)|>
  mutate(
   national_poll = factor(if_else(state == "National", 1, 0)),
      recency = as.numeric(difftime(as.Date("2024-11-05"), end_date, units = "days")),
    sample_size_weight = pmin(sample_size / 2300, 1),
    recency_weight = exp(-recency * 0.1),  
    combined_weight = recency_weight * sample_size_weight,
    candidate="Donald Trump"
  )
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| fig-cap: "Harris Kamala leads Donald Trump an average of 1% in Predicted Support in the 2024 US Election (From August 1, 2024, to November 2, 2024)."
#| fig-subcap: ["Unweighted","Weighted"]
#| layout-ncol: 1
#| label: fig-support-trend
# Plotting with the filtered data
plan(multisession)
##unweighted
ggplot(combined_data_total,aes(x = end_date,y=pct, colour = candidate)) +
  geom_point(alpha = 0.1)+ 
  geom_smooth(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted), method = "lm",size = 1)+ 
  geom_line(data = combined_predictions_filtered,aes(x=end_date,y = predicted_unweighted),size = 0.2)+
  labs(
       x = "End Date of the Poll",  # Change x-axis label to End Date
       y = "Predicted Support (%)",
       color = "Candidate",caption = 
                   "Model: Candidate Support Percentage = national_poll + pollster + population + state") +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),strip.background = element_rect(colour = "black", fill = "white")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week", limits = c(recent_date_cutoff, max_end_date)) +
  annotate("text", x = max(combined_predictions_filtered$end_date) -0.5, y = 48, 
           label = "47.5%", color = "blue", hjust = -0.2, size = 2) +  # Adjust size here
  annotate("text", x = max(combined_predictions_filtered$end_date) - 0.5, y = 46, 
           label = "46.5%", color = "red", hjust = -0.2, size = 2)
